---
title: "TimeSeries2"
author: "Andrew Mudie"
date: "2026-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(lubridate)
library(forecast)

set.seed(123)
```


```{r}

# ------------------------------------------------------------
# 1. Load the CSV
# ------------------------------------------------------------

csv_path = "C:/Users/andre/source/repos/ConflictTracker2/ConflictTracker/ConflictCalc/R/Sudan_Test2000_30July2026_Preprocessed.csv"
df = read.csv(csv_path, stringsAsFactors = FALSE)

# Inspect
head(df)
#head(df$periodStart, 20)
```



```{r}
# ------------------------------------------------------------
# 2. Convert periodStart to Date
# ------------------------------------------------------------

df$periodStart = as.Date(trimws(df$periodStart), format = "%d/%m/%Y")
# Inspect
head(df)
```

```{r}
#Add a month and season feature, based off the periodStart:
library(lubridate)
library(dplyr)

df <- df %>%
  mutate(
    month = month(periodStart),  # numeric month 1–12

    season = case_when(
      month %in% c(12, 1, 2) ~ "winter",
      month %in% c(3, 4, 5)  ~ "spring",
      month %in% c(6, 7, 8)  ~ "summer",
      month %in% c(9, 10, 11) ~ "autumn"
    ),

    season = factor(season, levels = c("winter", "spring", "summer", "autumn"))
  )

head(df)

```


```{r}
# ------------------------------------------------------------
# 3. Remove rows after July 2025: remove incomplete periods.
# ------------------------------------------------------------
cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
cutoff = as.Date("2025/07/01")
df = df %>% filter(periodStart <= cutoff)
cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```
```{r}

df <- df %>%
  arrange(country, place, periodStart) %>%   # correct ordering
  group_by(country, place) %>%
  mutate(
    # Local severity lags 2 and 3
    LocalEventCount_lag1 = lag(LocalEventCount, 1),
    LocalUniqueActorCount_lag1 = lag(LocalUniqueActorCount, 1),
    LocalDistinctEventTypes_lag1 = lag(LocalDistinctEventTypes, 1),
    LocalDistinctEventSubtypes_lag1 = lag(LocalDistinctEventSubtypes, 1),
    LocalTotalFatalities_lag1 = lag(LocalTotalFatalities, 1),
    LocalAvgSeverity_lag1 = lag(LocalAvgSeverity, 1),
    
    LocalEventCount_lag2 = lag(LocalEventCount, 2),
    LocalUniqueActorCount_lag2 = lag(LocalUniqueActorCount, 2),
    LocalDistinctEventTypes_lag2 = lag(LocalDistinctEventTypes, 2),
    LocalDistinctEventSubtypes_lag2 = lag(LocalDistinctEventSubtypes, 2),
    LocalTotalFatalities_lag2 = lag(LocalTotalFatalities, 2),
    LocalAvgSeverity_lag2 = lag(LocalAvgSeverity, 2),
      
    LocalEventCount_lag3 = lag(LocalEventCount, 3),
    LocalUniqueActorCount_lag3 = lag(LocalUniqueActorCount, 3),
    LocalDistinctEventTypes_lag3 = lag(LocalDistinctEventTypes, 3),
    LocalDistinctEventSubtypes_lag3 = lag(LocalDistinctEventSubtypes, 3),
    LocalTotalFatalities_lag3 = lag(LocalTotalFatalities, 3),
    LocalAvgSeverity_lag3 = lag(LocalAvgSeverity, 3),

    # Regional severity lags 1, 2 and 3

        RegionalEventCount_lag1 = lag(RegionalEventCount, 1),
    RegionalActorCount_lag1 = lag(RegionalActorCount, 1),
    RegionalDistinctEventSubTypes_lag1 = lag(RegionalDistinctEventSubTypes, 1),
    RegionalTotalFatalities_lag1 = lag(RegionalTotalFatalities, 1),
    RegionalAvgSeverity_lag1 = lag(RegionalAvgSeverity, 1),
    
    RegionalEventCount_lag2 = lag(RegionalEventCount, 2),
    RegionalActorCount_lag2 = lag(RegionalActorCount, 2),
    RegionalDistinctEventSubTypes_lag2 = lag(RegionalDistinctEventSubTypes, 2),
    RegionalTotalFatalities_lag2 = lag(RegionalTotalFatalities, 2),
    RegionalAvgSeverity_lag2 = lag(RegionalAvgSeverity, 2),

    RegionalEventCount_lag3 = lag(RegionalEventCount, 3),
    RegionalActorCount_lag3 = lag(RegionalActorCount, 3),
    RegionalDistinctEventSubTypes_lag3 = lag(RegionalDistinctEventSubTypes, 3),
    RegionalTotalFatalities_lag3 = lag(RegionalTotalFatalities, 3),
    RegionalAvgSeverity_lag3 = lag(RegionalAvgSeverity, 3),
    
    #Lags 1-3 for Regional actor types
RegionalProtestersEventCount_lag1 = lag(RegionalProtestersEventCount, 1),
RegionalStateForcesEventCount_lag1 = lag(RegionalStateForcesEventCount, 1),
RegionalPoliticalMilitiaEventCount_lag1 = lag(RegionalPoliticalMilitiaEventCount, 1),
RegionalIdentityMilitiaEventCount_lag1 = lag(RegionalIdentityMilitiaEventCount, 1),
RegionalRebelGroupEventCount_lag1 = lag(RegionalRebelGroupEventCount, 1),
RegionalCiviliansEventCount_lag_1 =lag(RegionalCiviliansEventCount, 1),
RegionalRiotersEventCount_lag1 = lag(RegionalRiotersEventCount, 1),
RegionalOtherTypeEventCount_lag1 = lag(RegionalOtherTypeEventCount, 1),
RegionalProtestersEventCount_lag2 = lag(RegionalProtestersEventCount, 2),
RegionalStateForcesEventCount_lag2 = lag(RegionalStateForcesEventCount, 2),
RegionalPoliticalMilitiaEventCount_lag2 = lag(RegionalPoliticalMilitiaEventCount, 2),
RegionalIdentityMilitiaEventCount_lag2 = lag(RegionalIdentityMilitiaEventCount, 2),
RegionalRebelGroupEventCount_lag2 = lag(RegionalRebelGroupEventCount, 2),
RegionalCiviliansEventCount_lag_2 =lag(RegionalCiviliansEventCount, 2),
RegionalRiotersEventCount_lag2 = lag(RegionalRiotersEventCount, 2),
RegionalOtherTypeEventCount_lag2 = lag(RegionalOtherTypeEventCount, 2),
RegionalProtestersEventCount_lag3 = lag(RegionalProtestersEventCount, 3),
RegionalStateForcesEventCount_lag3 = lag(RegionalStateForcesEventCount, 3),
RegionalPoliticalMilitiaEventCount_lag3 = lag(RegionalPoliticalMilitiaEventCount, 3),
RegionalIdentityMilitiaEventCount_lag3 = lag(RegionalIdentityMilitiaEventCount, 3),
RegionalRebelGroupEventCount_lag3 = lag(RegionalRebelGroupEventCount, 3),
RegionalCiviliansEventCount_lag_3 =lag(RegionalCiviliansEventCount, 3),
RegionalRiotersEventCount_lag3 = lag(RegionalRiotersEventCount, 3),
RegionalOtherTypeEventCount_lag3 = lag(RegionalOtherTypeEventCount, 3)
    
  ) %>%
  ungroup()
```

```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) local features
#"LocalEventCount","LocalUniqueActorCount","LocalDistinctEventTypes","LocalDistinctEventSubtypes","LocalTotalFatalities","LocalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_roll3 = rollapply(
      LocalEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_roll3 = rollapply(
      LocalUniqueActorCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_roll3 = rollapply(
      LocalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_roll3 = rollapply(
     LocalDistinctEventSubtypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_roll3 = rollapply(
     LocalTotalFatalities,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_roll3 = rollapply(
      LocalAvgSeverity,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Local actor types:
        LocalProtestersEventCount_roll3 = rollapply(
      LocalProtestersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        LocalStateForcesEventCount_roll3 = rollapply(
      LocalStateForcesEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_roll3 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_roll3 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_roll3 = rollapply(
      LocalRebelGroupEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_roll3 = rollapply(
      LocalCiviliansEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_roll3 = rollapply(
      LocalRiotersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_roll3 = rollapply(
      LocalOtherTypeEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  )
    
  ) %>%
  ungroup()
```

```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) regional features
#RegionalEventCount","RegionalActorCount","RegionalDistinctEventTypes","RegionalDistinctEventSubTypes","RegionalTotalFatalities","RegionalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    RegionalEventCount_roll3 = rollapply(
      RegionalEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
        RegionalActorCount_roll3 = rollapply(
     RegionalActorCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_roll3 = rollapply(
      RegionalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventTypes_roll3 = rollapply(
     RegionalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_roll3 = rollapply(
      RegionalTotalFatalities,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalAvgSeverity_roll3 = rollapply(
      RegionalAvgSeverity,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Rolling windows for Regional actor types:
    RegionalProtestersEventCount_roll3 = rollapply(
      RegionalProtestersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        RegionalStateForcesEventCount_roll3 = rollapply(
      RegionalStateForcesEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_roll3 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_roll3 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_roll3 = rollapply(
      RegionalRebelGroupEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_roll3 = rollapply(
      RegionalCiviliansEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_roll3 = rollapply(
      RegionalRiotersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_roll3 = rollapply(
      RegionalOtherTypeEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```

```{r}
head(df)
```
```{r}
#Add the volatility features (..._vol3, for both local and regional features):

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_vol3 = rollapply(
      LocalEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_vol3 = rollapply(
      LocalUniqueActorCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_vol3 = rollapply(
      LocalDistinctEventTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_vol3 = rollapply(
     LocalDistinctEventSubtypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_vol3 = rollapply(
     LocalTotalFatalities,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_vol3 = rollapply(
      LocalAvgSeverity,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
            RegionalEventCount_vol3 = rollapply(
      RegionalEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),

        RegionalActorCount_vol3 = rollapply(
     RegionalActorCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_vol3 = rollapply(
      RegionalDistinctEventTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventSubTypes_vol3 = rollapply(
     RegionalDistinctEventSubTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_vol3 = rollapply(
      RegionalTotalFatalities,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
      RegionalAvgSeverity_vol3 = rollapply(
      RegionalAvgSeverity,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  #Local actor types:
      LocalProtestersEventCount_vol3 = rollapply(
      LocalProtestersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 LocalStateForcesEventCount_vol3 = rollapply(
      LocalStateForcesEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_vol3 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_vol3 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_vol3 = rollapply(
      LocalRebelGroupEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_vol3 = rollapply(
      LocalCiviliansEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_vol3 = rollapply(
      LocalRiotersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_vol3 = rollapply(
      LocalOtherTypeEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
    
    
    
    
    
    #Regional actor types:
    RegionalProtestersEventCount_vol3 = rollapply(
      RegionalProtestersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 RegionalStateForcesEventCount_vol3 = rollapply(
      RegionalStateForcesEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_vol3 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_vol3 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_vol3 = rollapply(
      RegionalRebelGroupEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_vol3 = rollapply(
      RegionalCiviliansEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_vol3 = rollapply(
      RegionalRiotersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_vol3 = rollapply(
      RegionalOtherTypeEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```


```{r}
# ------------------------------------------------------------
# 4. Choose a target variable for forecasting
# ------------------------------------------------------------


target_var = "LocalAvgSeverity"   # set this to any column you want

target_colname = paste0("Target_", target_var)

df <- df %>%
  arrange(country, place, periodStart) %>%   # global sort for safety
  group_by(country, place) %>%              # correct grouping
  mutate(
    !!target_colname := lead(.data[[target_var]], 1)   # next period
  ) %>%
  ungroup()
```

```{r}
#Remove noise (places that are only present because something happened in the region)

cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
df <- df %>%
  group_by(country, place) %>%
  filter(sum(LocalEventCount, na.rm = TRUE) > 0) %>%
  ungroup()
cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```


```{r}

#write.csv(head(df), "df_columns_check.csv", row.names = FALSE)
#names(df)

#Remove duff columns (these are the CIA World Factbook columns for now, plus latitude and longitude and periodEnd)

cols_to_remove= c("latitude", "longitude", "periodEnd",
                  
   "LocalEventCount","LocalUniqueActorCount","LocalDistinctEventTypes","LocalDistinctEventSubtypes","LocalTotalFatalities","LocalAvgSeverity","RegionalEventCount","RegionalActorCount","RegionalDistinctEventTypes","RegionalDistinctEventSubTypes","RegionalTotalFatalities","RegionalAvgSeverity",              
                  
   
   "LocalProtestersEventCount", "LocalStateForcesEventCount", "LocalPoliticalMilitiaEventCount", "LocalCiviliansEventCount",   "LocalIdentityMilitiaEventCount",   "LocalRebelGroupEventCount",      "LocalRiotersEventCount",   "LocalOtherTypeEventCount",
   
      "RegionalProtestersEventCount", "RegionalStateForcesEventCount", "RegionalPoliticalMilitiaEventCount", "RegionalCiviliansEventCount",   "RegionalIdentityMilitiaEventCount",   "RegionalRebelGroupEventCount",      "RegionalRiotersEventCount",   "RegionalOtherTypeEventCount",
   
"Budget","Budget.surplus.....or.deficit....","Debt...external","Disputes...international","Ethnic.groups","Executive.branch","GDP..official.exchange.rate.","Government.type","Independence","International.law.organization.participation","International.organization.participation","Judicial.branch","Land.boundaries","Languages","Legal.system","Legislative.branch","Maritime.claims","Military.and.security.forces","Military.and.security.service.personnel.strengths","Military.deployments","Military.equipment.inventories.and.acquisitions","Military.expenditures","Military.service.age.and.obligation","Natural.resources","Political.parties","Population","Population.below.poverty.line","Population.distribution","Population.growth.rate","Public.debt","Real.GDP..purchasing.power.parity.","Real.GDP.growth.rate","Real.GDP.per.capita","Refugees.and.internally.displaced.persons","Religions","Terrain","Trafficking.in.persons","Unemployment.rate","Urbanization","Youth.unemployment.rate..ages.15.24."
)



# Iterate and drop columns one by one
for (col in cols_to_remove) {
  if (col %in% names(df)) {
    df = df %>% select(-all_of(col))
  }
}


names(df)
summary(df)

```




```{r}
# ------------------------------------------------------------
# 5. Create numeric time index and split into train/test
# ------------------------------------------------------------

# Ensure periodStart is a Date
df$periodStart = as.Date(df$periodStart)

# Compute the minimum date across the full dataset
min_date = min(df$periodStart, na.rm = TRUE)

# Create numeric DaysSinceMinPeriodStart
df = df %>%
  mutate(DaysSinceMinPeriodStart = as.numeric(periodStart - min_date))

# ------------------------------------------------------------
# Remove earliest period per place (lag features invalid there)
# ------------------------------------------------------------

earliest = df %>%
  group_by(place) %>%
  summarise(earliest_period = min(periodStart)) %>%
  ungroup()

df_no_earliest = df %>%
  left_join(earliest, by = "place") %>%
  filter(periodStart != earliest_period)

# ------------------------------------------------------------
# Define cutoff dates
# ------------------------------------------------------------

T1 = as.Date("2025-01-22")
T2 = as.Date("2025-05-14")   # last complete period

# ------------------------------------------------------------
# Split into training and test sets
# ------------------------------------------------------------

train_df = df_no_earliest %>%
  filter(periodStart <= T1) %>%
  select(-periodStart)   # drop original date column

test_df = df_no_earliest %>%
  filter(periodStart > T1 & periodStart <= T2) %>%
  select(-periodStart)   # drop original date column

# Optional: capture earliest rows separately
#earliest_rows = df %>%
 # inner_join(earliest, by = "place") %>%
 # filter(periodStart == earliest_period)

train_df$earliest_period = NULL
test_df$earliest_period = NULL

head(train_df)
head(test_df)

write.csv(train_df, "train_df.csv", row.names = FALSE)
write.csv(test_df,  "test_df.csv",  row.names = FALSE)
```


```{r}
#Install Catboost (requires rlang 1.2.0+ and rtools 4.5+)


#install.packages("devtools") 
devtools::install_github("catboost/catboost", subdir = "catboost/R-package")
library(catboost)

library(dplyr)
```

```{r}
# ------------------------------------------------------------
# 6. train a catboost model based on country and place forming the composite key. 
# ------------------------------------------------------------


head(train_df)

#Prepare CatBoost training pool

print(cat("Target column: ", target_colname))

cat_features = c("country", "place")

train_pool = catboost.load_pool(
  data  = train_df %>% select(-all_of(target_colname)),
  label = train_df[[target_colname]],
  cat_features = cat_features
)

#Train the global CatBoost model
params = list(
  loss_function = "RMSE",
  iterations = 500,
  depth = 8,
  learning_rate = 0.05
)

model_catBoost = catboost.train(train_pool, params = params)

```



```{r}
# ------------------------------------------------------------
# 4. Compute training evaluation metrics
# ------------------------------------------------------------

test_pool <- catboost.load_pool(
  data  = test_df %>% select(-all_of(target_colname)),
  label = test_df[[target_colname]],
  cat_features = cat_features
)

# ------------------------------------------------------------
# Generate predictions
# ------------------------------------------------------------

preds <- catboost.predict(model_catBoost, test_pool)

# Actual values
actuals <- test_df[[target_colname]]

# ------------------------------------------------------------
# Evaluation metrics
# ------------------------------------------------------------

# RMSE
rmse <- sqrt(mean((preds - actuals)^2))

# MAE
mae <- mean(abs(preds - actuals))

# MAPE (avoid division by zero)
mape <- mean(abs((preds - actuals) / pmax(actuals, 1e-8))) * 100

# R-squared
ss_res <- sum((actuals - preds)^2)
ss_tot <- sum((actuals - mean(actuals))^2)
r2 <- 1 - ss_res/ss_tot

# ------------------------------------------------------------
# Print results
# ------------------------------------------------------------

cat("RMSE: ", rmse, "\n")
cat("MAE:  ", mae, "\n")
cat("MAPE: (not suitable if 0s in dataset)", mape, "%\n")
cat("R²:   ", r2, "\n")

```
RMSE is much higher than MAE- making a few huge mistakes.
Low R^2 means the model is underfitting. 

A couple of options here. 
-Switch target metric- regional severity?
-Add more features (actor x/y/z active)
-Add seasonality (from month of startperiod)
-Remove raw local/regional features
-Add lag1/lag2/lag3
-Add rolling mean/std (3 periods)
-Add actor presence lag1
-Add Distance from Border


Update 28th July 2026.
-Added lag features (raw values at T-1, T-2 and T-3)
-Added trend features (mean over 3 prior periods)
-Added volatility features (std dev).
-Switched metric to RegionalAvgSeverity
-Fixed an issue with incorrect ordering
-Removed raw features.
-Added seasonal features season and month
-And still R^2 is terrible at 0.04! 

Next actions: Add some actor based features (Is_Rebel_Present, Is_State_Military_present, etc)
-Spatial features (distance from border, distance from capital)
-

