---
title: "R Notebook"
output: html_notebook
---

```{r}
train_df = read.csv("train_df.csv", stringsAsFactors = TRUE)
test_df = read.csv("test_df.csv", stringsAsFactors = TRUE)
```

```{r}

target_colname = "Target_ConflictFlag"

target <- target_colname
cat_features <- c("country", "place")

library(dplyr) #gives the %>% operator

# Convert categorical features to integer encoding
train_df_lgb <- train_df %>%
  mutate(across(all_of(cat_features), as.integer))

test_df_lgb <- test_df %>%
  mutate(across(all_of(cat_features), as.integer))

# Extract labels
train_label <- train_df_lgb[[target]]
test_label  <- test_df_lgb[[target]]

# Remove target column from features
train_matrix <- as.matrix(train_df_lgb %>% select(-all_of(target)))
test_matrix  <- as.matrix(test_df_lgb %>% select(-all_of(target)))

#Build LightGBM datasets
dtrain <- lgb.Dataset(
  data = train_matrix,
  label = train_label,
  categorical_feature = which(colnames(train_matrix) %in% cat_features)
)

dtest <- lgb.Dataset(
  data = test_matrix,
  label = test_label,
  categorical_feature = which(colnames(test_matrix) %in% cat_features)
)

params <- list(
  objective = "binary",
  metric = "auc",
  boosting = "gbdt",
  learning_rate = 0.05,
  num_leaves = 31,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq = 5
)

#Train LightGBM model (binary classification)
model_lgb <- lgb.train(
  params = params,
  data = dtrain,
  nrounds = 500,
  valids = list(test = dtest),
  verbose = 1
)

lgb.save(model_lgb, "model_lgb.txt")

```

# Evaluation of Model
```{r}
pred_probs <- predict(model_lgb, test_matrix)

threshold <- 0.5
pred_classes <- ifelse(pred_probs >= threshold, 1, 0)

# AUC
auc <- pROC::auc(test_label, pred_probs)

# Accuracy
accuracy <- mean(pred_classes == test_label)

# Precision
precision <- sum(pred_classes == 1 & test_label == 1) /
             sum(pred_classes == 1)

# Recall
recall <- sum(pred_classes == 1 & test_label == 1) /
          sum(test_label == 1)

# F1 Score
f1 <- 2 * (precision * recall) / (precision + recall)

# Confusion Matrix
conf_matrix <- table(Predicted = pred_classes, Actual = test_label)

cat("AUC:        ", auc, "\n")
cat("Accuracy:   ", accuracy, "\n")
cat("Precision:  ", precision, "\n")
cat("Recall:     ", recall, "\n")
cat("F1 Score:   ", f1, "\n")
```



```{r}
# ============================
# 1. Load libraries
# ============================
library(lightgbm)
library(data.table)
library(shapviz)

# ============================
# 2. Load training data from CSV
# ============================
X_train <- fread("train_df.csv")

# Convert to data.frame if needed
X_train <- as.data.frame(X_train)


# ============================
# 3. Load LightGBM model
# ============================
# If you saved using lgb.save()
model_lgb <- lgb.load("model_lgb.txt")

# OR if you saved using saveRDS()
# model_lgb <- readRDS("model_lgb.rds")

# ============================
# 4. Compute TRUE SHAP values
# ============================
#sv <- shapviz(model_lgb,  as.matrix(X_train))

set.seed(123)
X_sample <- train_matrix[sample(nrow(train_matrix), 50000), ]
sv <- shapviz(model_lgb, X_sample)

# ============================
# 5. Display SHAP importance
# ============================
# Global importance table
sv_importance(sv)

# Bar plot of SHAP importance
sv_importance(sv, kind = "bar")

# Beeswarm plot (best global view)
sv_importance(sv, kind = "beeswarm")

# ============================
# 6. SHAP dependence plot
# ============================
# Replace with any feature name you want to inspect
sv_dependence(sv, "LocalTotalFatalities_roll12")

# ============================
# 7. Access raw SHAP values
# ============================
shap_values <- sv$S
head(shap_values)

# ============================
# 8. Baseline value (average model prediction)
# ============================
sv$baseline

```
Mean shap is a neat way of understanding which features push the prediction. We can see the most influential feature is RegionalAvgSeverity_roll12: the rolling mean of severity over the prior 12 periods. Country and Place are also influential- some places appear to be inherently more conflict prone than others, perhaps this is expected given conflict breeds conflict. 12 and 3 periods dominate the notable features: again supporting the idea that long term conflict breeds conflict, but ConflictFlag is also notable: current presence of conflice is a notable indictor of future conflict. The only actor type feature that is shown is RegionalRevelGroupEventCount_roll12: the sustained presence of rebel groups within the region in the last 12 breeds conflict. Days since min period start also is influential, suggesting there is a baseline level of conflict increasing as time does, and this is supported in the more imbalanced training set (the end of the period).

# Calculate P value. 
```{r}
p <- 1 / (1 + exp(sv$baseline * -1))
p
```
Before features add any value, we have a P value 0.78. This means a 78% chance of conflict in the next six months before taking into account features. As expected, as the dataset is imbalanced. 

# Imbalance Ratio
```{r}

print("Imbalance ratio of train_df:")
train_df %>%
  count(Target_ConflictFlag) %>%
  mutate(
    proportion = n / sum(n),
    imbalance_ratio = max(n) / min(n)
  )

print("Imbalance ratio of test_df:")
test_df %>%
  count(Target_ConflictFlag) %>%
  mutate(
    proportion = n / sum(n),
    imbalance_ratio = max(n) / min(n)
  )
```