---
title: "TimeSeries3"
author: "Andrew Mudie"
date: "2026-07-30"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(dplyr)
library(lubridate)
library(forecast)

set.seed(123)
```

# Load Raw Data
```{r}

# ------------------------------------------------------------
# 1. Load the CSV
# ------------------------------------------------------------

csv_path = "C:/Users/andre/source/repos/ConflictTracker2/ConflictTracker/ConflictCalc/R/eastafrica_20150101_5d2a4f05-0bef-4bc6-aee7-853dab0d5f7b.csv"
df = read.csv(csv_path, stringsAsFactors = FALSE)

# Inspect
head(df)
#head(df$periodStart, 20)
```


# Data Conversion
```{r}
# ------------------------------------------------------------
# 2. Convert periodStart to Date
# ------------------------------------------------------------

df$periodStart <- as.Date(trimws(df$periodStart), format = "%Y-%m-%d")
# Inspect
head(df)
```

# Add seasonal features (Month, Season)
```{r}
#Add a month and season feature, based off the periodStart:
library(lubridate)
library(dplyr)

df <- df %>%
  mutate(
    month = month(periodStart),  # numeric month 1–12

    season = case_when(
      month %in% c(12, 1, 2) ~ "winter",
      month %in% c(3, 4, 5)  ~ "spring",
      month %in% c(6, 7, 8)  ~ "summer",
      month %in% c(9, 10, 11) ~ "autumn"
    ),

    season = factor(season, levels = c("winter", "spring", "summer", "autumn"))
  )

head(df)

```
# Remove country places
```{r}
cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
df =  subset(df, country != place)
cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```

# Lag features from prior periods
```{r}
# ------------------------------------------------------------
# 3. Remove rows after July 2025: remove incomplete periods.
# ------------------------------------------------------------
cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
cutoff = as.Date("2025/07/01")
df = df %>% filter(periodStart <= cutoff)
cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```
```{r}

df <- df %>%
  arrange(country, place, periodStart) %>%   # correct ordering
  group_by(country, place) %>%
  mutate(
    # Local severity lags 1 - 3
    LocalEventCount_lag1 = lag(LocalEventCount, 1),
    LocalUniqueActorCount_lag1 = lag(LocalUniqueActorCount, 1),
    LocalDistinctEventTypes_lag1 = lag(LocalDistinctEventTypes, 1),
    LocalDistinctEventSubtypes_lag1 = lag(LocalDistinctEventSubtypes, 1),
    LocalTotalFatalities_lag1 = lag(LocalTotalFatalities, 1),
    LocalAvgSeverity_lag1 = lag(LocalAvgSeverity, 1),
    
    LocalEventCount_lag2 = lag(LocalEventCount, 2),
    LocalUniqueActorCount_lag2 = lag(LocalUniqueActorCount, 2),
    LocalDistinctEventTypes_lag2 = lag(LocalDistinctEventTypes, 2),
    LocalDistinctEventSubtypes_lag2 = lag(LocalDistinctEventSubtypes, 2),
    LocalTotalFatalities_lag2 = lag(LocalTotalFatalities, 2),
    LocalAvgSeverity_lag2 = lag(LocalAvgSeverity, 2),
      
    LocalEventCount_lag3 = lag(LocalEventCount, 3),
    LocalUniqueActorCount_lag3 = lag(LocalUniqueActorCount, 3),
    LocalDistinctEventTypes_lag3 = lag(LocalDistinctEventTypes, 3),
    LocalDistinctEventSubtypes_lag3 = lag(LocalDistinctEventSubtypes, 3),
    LocalTotalFatalities_lag3 = lag(LocalTotalFatalities, 3),
    LocalAvgSeverity_lag3 = lag(LocalAvgSeverity, 3),

    
        LocalEventCount_lag12 = lag(LocalEventCount, 12),
    LocalUniqueActorCount_lag12 = lag(LocalUniqueActorCount, 12),
    LocalDistinctEventTypes_lag12 = lag(LocalDistinctEventTypes, 12),
    LocalDistinctEventSubtypes_lag12 = lag(LocalDistinctEventSubtypes, 12),
    LocalTotalFatalities_lag12 = lag(LocalTotalFatalities, 12),
    LocalAvgSeverity_lag12 = lag(LocalAvgSeverity, 12),
    
    # Regional severity lags 1, 2 and 3

        RegionalEventCount_lag1 = lag(RegionalEventCount, 1),
    RegionalActorCount_lag1 = lag(RegionalActorCount, 1),
    RegionalDistinctEventSubTypes_lag1 = lag(RegionalDistinctEventSubTypes, 1),
    RegionalTotalFatalities_lag1 = lag(RegionalTotalFatalities, 1),
    RegionalAvgSeverity_lag1 = lag(RegionalAvgSeverity, 1),
    
    RegionalEventCount_lag2 = lag(RegionalEventCount, 2),
    RegionalActorCount_lag2 = lag(RegionalActorCount, 2),
    RegionalDistinctEventSubTypes_lag2 = lag(RegionalDistinctEventSubTypes, 2),
    RegionalTotalFatalities_lag2 = lag(RegionalTotalFatalities, 2),
    RegionalAvgSeverity_lag2 = lag(RegionalAvgSeverity, 2),

    RegionalEventCount_lag3 = lag(RegionalEventCount, 3),
    RegionalActorCount_lag3 = lag(RegionalActorCount, 3),
    RegionalDistinctEventSubTypes_lag3 = lag(RegionalDistinctEventSubTypes, 3),
    RegionalTotalFatalities_lag3 = lag(RegionalTotalFatalities, 3),
    RegionalAvgSeverity_lag3 = lag(RegionalAvgSeverity, 3),
    
        RegionalEventCount_lag12 = lag(RegionalEventCount, 12),
    RegionalActorCount_lag12 = lag(RegionalActorCount, 12),
    RegionalDistinctEventSubTypes_lag12 = lag(RegionalDistinctEventSubTypes, 12),
    RegionalTotalFatalities_lag12 = lag(RegionalTotalFatalities, 12),
    RegionalAvgSeverity_lag12 = lag(RegionalAvgSeverity, 12),
    
    #Lags 1-3 for Regional actor types
RegionalProtestersEventCount_lag1 = lag(RegionalProtestersEventCount, 1),
RegionalStateForcesEventCount_lag1 = lag(RegionalStateForcesEventCount, 1),
RegionalPoliticalMilitiaEventCount_lag1 = lag(RegionalPoliticalMilitiaEventCount, 1),
RegionalIdentityMilitiaEventCount_lag1 = lag(RegionalIdentityMilitiaEventCount, 1),
RegionalRebelGroupEventCount_lag1 = lag(RegionalRebelGroupEventCount, 1),
RegionalCiviliansEventCount_lag_1 =lag(RegionalCiviliansEventCount, 1),
RegionalRiotersEventCount_lag1 = lag(RegionalRiotersEventCount, 1),
RegionalOtherTypeEventCount_lag1 = lag(RegionalOtherTypeEventCount, 1),
RegionalProtestersEventCount_lag2 = lag(RegionalProtestersEventCount, 2),
RegionalStateForcesEventCount_lag2 = lag(RegionalStateForcesEventCount, 2),
RegionalPoliticalMilitiaEventCount_lag2 = lag(RegionalPoliticalMilitiaEventCount, 2),
RegionalIdentityMilitiaEventCount_lag2 = lag(RegionalIdentityMilitiaEventCount, 2),
RegionalRebelGroupEventCount_lag2 = lag(RegionalRebelGroupEventCount, 2),
RegionalCiviliansEventCount_lag_2 =lag(RegionalCiviliansEventCount, 2),
RegionalRiotersEventCount_lag2 = lag(RegionalRiotersEventCount, 2),
RegionalOtherTypeEventCount_lag2 = lag(RegionalOtherTypeEventCount, 2),
RegionalProtestersEventCount_lag3 = lag(RegionalProtestersEventCount, 3),
RegionalStateForcesEventCount_lag3 = lag(RegionalStateForcesEventCount, 3),
RegionalPoliticalMilitiaEventCount_lag3 = lag(RegionalPoliticalMilitiaEventCount, 3),
RegionalIdentityMilitiaEventCount_lag3 = lag(RegionalIdentityMilitiaEventCount, 3),
RegionalRebelGroupEventCount_lag3 = lag(RegionalRebelGroupEventCount, 3),
RegionalCiviliansEventCount_lag_3 =lag(RegionalCiviliansEventCount, 3),
RegionalRiotersEventCount_lag3 = lag(RegionalRiotersEventCount, 3),
RegionalOtherTypeEventCount_lag3 = lag(RegionalOtherTypeEventCount, 3),

RegionalProtestersEventCount_lag12 = lag(RegionalProtestersEventCount, 12),
RegionalStateForcesEventCount_lag12 = lag(RegionalStateForcesEventCount, 12),
RegionalPoliticalMilitiaEventCount_lag12 = lag(RegionalPoliticalMilitiaEventCount, 12),
RegionalIdentityMilitiaEventCount_lag12 = lag(RegionalIdentityMilitiaEventCount, 12),
RegionalRebelGroupEventCount_lag12 = lag(RegionalRebelGroupEventCount, 12),
RegionalCiviliansEventCount_lag_12 =lag(RegionalCiviliansEventCount, 12),
RegionalRiotersEventCount_lag12 = lag(RegionalRiotersEventCount, 12),
RegionalOtherTypeEventCount_lag12 = lag(RegionalOtherTypeEventCount, 12)

    
  ) %>%
  ungroup()
```


# Local rolling means over three features
```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) local features
#"LocalEventCount","LocalUniqueActorCount","LocalDistinctEventTypes","LocalDistinctEventSubtypes","LocalTotalFatalities","LocalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_roll3 = rollapply(
      LocalEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_roll3 = rollapply(
      LocalUniqueActorCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_roll3 = rollapply(
      LocalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_roll3 = rollapply(
     LocalDistinctEventSubtypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_roll3 = rollapply(
     LocalTotalFatalities,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_roll3 = rollapply(
      LocalAvgSeverity,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Local actor types:
        LocalProtestersEventCount_roll3 = rollapply(
      LocalProtestersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        LocalStateForcesEventCount_roll3 = rollapply(
      LocalStateForcesEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_roll3 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_roll3 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_roll3 = rollapply(
      LocalRebelGroupEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_roll3 = rollapply(
      LocalCiviliansEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_roll3 = rollapply(
      LocalRiotersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_roll3 = rollapply(
      LocalOtherTypeEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  )
    
  ) %>%
  ungroup()
```

# Regional rolling means (3 periods)
```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) regional features
#RegionalEventCount","RegionalActorCount","RegionalDistinctEventTypes","RegionalDistinctEventSubTypes","RegionalTotalFatalities","RegionalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    RegionalEventCount_roll3 = rollapply(
      RegionalEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
        RegionalActorCount_roll3 = rollapply(
     RegionalActorCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_roll3 = rollapply(
      RegionalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventTypes_roll3 = rollapply(
     RegionalDistinctEventTypes,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_roll3 = rollapply(
      RegionalTotalFatalities,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalAvgSeverity_roll3 = rollapply(
      RegionalAvgSeverity,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Rolling windows for Regional actor types:
    RegionalProtestersEventCount_roll3 = rollapply(
      RegionalProtestersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        RegionalStateForcesEventCount_roll3 = rollapply(
      RegionalStateForcesEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_roll3 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_roll3 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_roll3 = rollapply(
      RegionalRebelGroupEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_roll3 = rollapply(
      RegionalCiviliansEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_roll3 = rollapply(
      RegionalRiotersEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_roll3 = rollapply(
      RegionalOtherTypeEventCount,
      width = 3,
      FUN = mean,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```

# Local rolling means over three features
```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) local features
#"LocalEventCount","LocalUniqueActorCount","LocalDistinctEventTypes","LocalDistinctEventSubtypes","LocalTotalFatalities","LocalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_roll12 = rollapply(
      LocalEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_roll12 = rollapply(
      LocalUniqueActorCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_roll12 = rollapply(
      LocalDistinctEventTypes,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_roll12 = rollapply(
     LocalDistinctEventSubtypes,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_roll12 = rollapply(
     LocalTotalFatalities,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_roll12 = rollapply(
      LocalAvgSeverity,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Local actor types:
        LocalProtestersEventCount_roll12 = rollapply(
      LocalProtestersEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        LocalStateForcesEventCount_roll12 = rollapply(
      LocalStateForcesEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_roll12 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_roll12 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_roll12 = rollapply(
      LocalRebelGroupEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_roll12 = rollapply(
      LocalCiviliansEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_roll12 = rollapply(
      LocalRiotersEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_roll12 = rollapply(
      LocalOtherTypeEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  )
    
  ) %>%
  ungroup()
```

# Regional rolling means (12 periods)
```{r}
# ------------------------------------------------------------
# Add longer trend (avg over previous three periods) regional features
#RegionalEventCount","RegionalActorCount","RegionalDistinctEventTypes","RegionalDistinctEventSubTypes","RegionalTotalFatalities","RegionalAvgSeverity"
# ------------------------------------------------------------

library(dplyr)
library(zoo)

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    RegionalEventCount_roll12 = rollapply(
      RegionalEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
        RegionalActorCount_roll12 = rollapply(
     RegionalActorCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_roll12 = rollapply(
      RegionalDistinctEventTypes,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventTypes_roll12 = rollapply(
     RegionalDistinctEventTypes,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_roll12 = rollapply(
      RegionalTotalFatalities,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
        RegionalAvgSeverity_roll12 = rollapply(
      RegionalAvgSeverity,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
    ),
    
    #Rolling windows for Regional actor types:
    RegionalProtestersEventCount_roll12 = rollapply(
      RegionalProtestersEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	        RegionalStateForcesEventCount_roll12 = rollapply(
      RegionalStateForcesEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_roll12 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_roll12 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_roll12 = rollapply(
      RegionalRebelGroupEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_roll12 = rollapply(
      RegionalCiviliansEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_roll12 = rollapply(
      RegionalRiotersEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_roll12 = rollapply(
      RegionalOtherTypeEventCount,
      width = 12,
      FUN = mean,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```

#Local and regional volatility features (Standard deviation over 3 periods):
```{r}
#Add the volatility features (..._vol3, for both local and regional features):

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_vol3 = rollapply(
      LocalEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_vol3 = rollapply(
      LocalUniqueActorCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_vol3 = rollapply(
      LocalDistinctEventTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_vol3 = rollapply(
     LocalDistinctEventSubtypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_vol3 = rollapply(
     LocalTotalFatalities,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_vol3 = rollapply(
      LocalAvgSeverity,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
            RegionalEventCount_vol3 = rollapply(
      RegionalEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),

        RegionalActorCount_vol3 = rollapply(
     RegionalActorCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_vol3 = rollapply(
      RegionalDistinctEventTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventSubTypes_vol3 = rollapply(
     RegionalDistinctEventSubTypes,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_vol3 = rollapply(
      RegionalTotalFatalities,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
      RegionalAvgSeverity_vol3 = rollapply(
      RegionalAvgSeverity,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  #Local actor types:
      LocalProtestersEventCount_vol3 = rollapply(
      LocalProtestersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 LocalStateForcesEventCount_vol3 = rollapply(
      LocalStateForcesEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_vol3 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_vol3 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_vol3 = rollapply(
      LocalRebelGroupEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_vol3 = rollapply(
      LocalCiviliansEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_vol3 = rollapply(
      LocalRiotersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_vol3 = rollapply(
      LocalOtherTypeEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
    
    
    
    
    
    #Regional actor types:
    RegionalProtestersEventCount_vol3 = rollapply(
      RegionalProtestersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 RegionalStateForcesEventCount_vol3 = rollapply(
      RegionalStateForcesEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_vol3 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_vol3 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_vol3 = rollapply(
      RegionalRebelGroupEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_vol3 = rollapply(
      RegionalCiviliansEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_vol3 = rollapply(
      RegionalRiotersEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_vol3 = rollapply(
      RegionalOtherTypeEventCount,
      width = 3,
      FUN = sd,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```

```{r}
#Add the volatility features (..._vol12, for both local and regional features):

df <- df %>%
  arrange(country, place, periodStart) %>%   # ensure correct ordering
  group_by(place, country) %>%
  mutate(
    LocalEventCount_vol12 = rollapply(
      LocalEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       LocalUniqueActorCount_vol12 = rollapply(
      LocalUniqueActorCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventTypes_vol12 = rollapply(
      LocalDistinctEventTypes,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalDistinctEventSubtypes_vol12 = rollapply(
     LocalDistinctEventSubtypes,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalTotalFatalities_vol12 = rollapply(
     LocalTotalFatalities,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        LocalAvgSeverity_vol12 = rollapply(
      LocalAvgSeverity,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
            RegionalEventCount_vol12 = rollapply(
      RegionalEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),

        RegionalActorCount_vol12 = rollapply(
     RegionalActorCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  
    
        RegionalDistinctEventTypes_vol12 = rollapply(
      RegionalDistinctEventTypes,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
       RegionalDistinctEventSubTypes_vol12 = rollapply(
     RegionalDistinctEventSubTypes,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
        RegionalTotalFatalities_vol12 = rollapply(
      RegionalTotalFatalities,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
    
      RegionalAvgSeverity_vol12 = rollapply(
      RegionalAvgSeverity,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
    ),
  #Local actor types:
      LocalProtestersEventCount_vol12 = rollapply(
      LocalProtestersEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 LocalStateForcesEventCount_vol12 = rollapply(
      LocalStateForcesEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalPoliticalMilitiaEventCount_vol12 = rollapply(
      LocalPoliticalMilitiaEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalIdentityMilitiaEventCount_vol12 = rollapply(
      LocalIdentityMilitiaEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   LocalRebelGroupEventCount_vol12 = rollapply(
      LocalRebelGroupEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalCiviliansEventCount_vol12 = rollapply(
      LocalCiviliansEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalRiotersEventCount_vol12 = rollapply(
      LocalRiotersEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  LocalOtherTypeEventCount_vol12 = rollapply(
      LocalOtherTypeEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
    
    
    
    
    
    #Regional actor types:
    RegionalProtestersEventCount_vol12 = rollapply(
      RegionalProtestersEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  

	 RegionalStateForcesEventCount_vol12 = rollapply(
      RegionalStateForcesEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalPoliticalMilitiaEventCount_vol12 = rollapply(
      RegionalPoliticalMilitiaEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalIdentityMilitiaEventCount_vol12 = rollapply(
      RegionalIdentityMilitiaEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	   RegionalRebelGroupEventCount_vol12 = rollapply(
      RegionalRebelGroupEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalCiviliansEventCount_vol12 = rollapply(
      RegionalCiviliansEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalRiotersEventCount_vol12 = rollapply(
      RegionalRiotersEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  ),
	  
	  RegionalOtherTypeEventCount_vol12 = rollapply(
      RegionalOtherTypeEventCount,
      width = 12,
      FUN = sd,
      align = "right",
      fill = NA
	  )
  ) %>%
  ungroup()
```


# End of adding time series features
```{r}
#Finished adding the lag, rolling avg and volatility features. 
head(df)
```

# Application of target variable
```{r, echo = FALSE}
# ------------------------------------------------------------
# 4. Choose a target variable for forecasting
# ------------------------------------------------------------

#Target feature is presence of conflice in T+1.
target_var <- "ConflictFlag"
target_colname <- paste0("Target_", target_var)

#df <- df %>%
#  arrange(country, place, periodStart) %>%
#  group_by(country, place) %>%
#  mutate(
 #   Target_ConflictFlag = lead(
#      as.integer(
#        (LocalAvgSeverity >= 3) |
#        (LocalTotalFatalities >= 3) |
 #       (RegionalAvgSeverity >= 3) |
#        (RegionalTotalFatalities >= 3)
#      ),
#      1
#    )
#  ) %>%
#  ungroup() #One month prediction only


#six months (periods):
df <- df %>%
  arrange(country, place, periodStart) %>%
  group_by(country, place) %>%
  mutate(
    ConflictFlag = as.integer(
      (LocalAvgSeverity >= 3) |
      (LocalTotalFatalities >= 3) |
      (RegionalAvgSeverity >= 3) |
      (RegionalTotalFatalities >= 3)
    ),
    Target_ConflictFlag = as.integer(
      Reduce(`|`, lapply(1:6, function(k) lead(ConflictFlag, k)))
    )
  ) %>%
  ungroup()


df = df %>%
  filter(!is.na(df[[target_colname]]))

```

# Removal of incomplete periods (T1, T2)
```{r}
#Remove any null ags (incomplete periods)
df <- df %>%
  filter(
    !is.na(LocalDistinctEventSubtypes_lag12),
    !is.na(RegionalDistinctEventSubTypes_lag12)
  )
head(df)
```



```{r}
#Remove noise (places that are only present because something happened in the region)

#cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
#df <- df %>%
#  group_by(country, place) %>%
#  filter(sum(LocalEventCount, na.rm = TRUE) > 0) %>%
#  ungroup()
#cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```


```{r}
#Remove rows for which there are 0 events for the periodStart. They will be 0 if there is no event data for that period. 
#cat("Before removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
#df <- df %>%
#  group_by(periodStart) %>%
#  filter(any(LocalEventCount > 0)) %>%
#  ungroup()
#cat("After removing Rows:", nrow(df), "  Columns:", ncol(df), "\n")
```

```{r}
#Summarise:
df %>%
  summarise(
    distinct_periods = n_distinct(periodStart),
    min_period = min(periodStart, na.rm = TRUE),
    max_period = max(periodStart, na.rm = TRUE)
  )

df %>%
  distinct(periodStart) %>%
  arrange(periodStart)
```

```{r}
write.csv(df,  "df.csv",  row.names = FALSE)
```


# Removal of unneeded columns
```{r}

cols_to_remove= c("latitude", "longitude", "periodEnd",
                  
   "LocalEventCount","LocalUniqueActorCount","LocalDistinctEventTypes","LocalDistinctEventSubtypes","LocalTotalFatalities","LocalAvgSeverity","RegionalEventCount","RegionalActorCount","RegionalDistinctEventTypes","RegionalDistinctEventSubTypes","RegionalTotalFatalities","RegionalAvgSeverity",              
                  
   
   "LocalProtestersEventCount", "LocalStateForcesEventCount", "LocalPoliticalMilitiaEventCount", "LocalCiviliansEventCount",   "LocalIdentityMilitiaEventCount",   "LocalRebelGroupEventCount",      "LocalRiotersEventCount",   "LocalOtherTypeEventCount",
   
      "RegionalProtestersEventCount", "RegionalStateForcesEventCount", "RegionalPoliticalMilitiaEventCount", "RegionalCiviliansEventCount",   "RegionalIdentityMilitiaEventCount",   "RegionalRebelGroupEventCount",      "RegionalRiotersEventCount",   "RegionalOtherTypeEventCount",
   
"Budget","Budget.surplus.....or.deficit....","Debt...external","Disputes...international","Ethnic.groups","Executive.branch","GDP..official.exchange.rate.","Government.type","Independence","International.law.organization.participation","International.organization.participation","Judicial.branch","Land.boundaries","Languages","Legal.system","Legislative.branch","Maritime.claims","Military.and.security.forces","Military.and.security.service.personnel.strengths","Military.deployments","Military.equipment.inventories.and.acquisitions","Military.expenditures","Military.service.age.and.obligation","Natural.resources","Political.parties","Population","Population.below.poverty.line","Population.distribution","Population.growth.rate","Public.debt","Real.GDP..purchasing.power.parity.","Real.GDP.growth.rate","Real.GDP.per.capita","Refugees.and.internally.displaced.persons","Religions","Terrain","Trafficking.in.persons","Unemployment.rate","Urbanization","Youth.unemployment.rate..ages.15.24."
)



# Iterate and drop columns one by one
for (col in cols_to_remove) {
  if (col %in% names(df)) {
    df = df %>% select(-all_of(col))
  }
}

names(df)
summary(df)

```



# Dataset splitting
## Identification of date ranges of testing and training data 
```{r}
# ------------------------------------------------------------
# 5. Create numeric time index and split into train/test
# ------------------------------------------------------------

# Ensure periodStart is a Date
df$periodStart = as.Date(df$periodStart)

# Compute the minimum date across the full dataset
min_date = min(df$periodStart, na.rm = TRUE)

# Create numeric DaysSinceMinPeriodStart
df = df %>%
  mutate(DaysSinceMinPeriodStart = as.numeric(periodStart - min_date))

# ------------------------------------------------------------
# Remove earliest period per place (lag features invalid there)
# ------------------------------------------------------------

earliest = df %>%
  group_by(place) %>%
  summarise(earliest_period = min(periodStart)) %>%
  ungroup()

df_no_earliest = df %>%
  left_join(earliest, by = "place") %>%
  filter(periodStart != earliest_period)

# ------------------------------------------------------------
# Define cutoff dates
# ------------------------------------------------------------

# Get sorted distinct periods
sorted_periods <- df %>%
  distinct(periodStart) %>%
  arrange(periodStart) %>%
  pull(periodStart)

periodsInTestData = 18


# Max period
min_period <- head(sorted_periods, 1)

max_period <- tail(sorted_periods, 1)

# Lag 12 period
T2 = sorted_periods[length(sorted_periods) - periodsInTestData]

# Lag 13 period (one step earlier)
T1 = sorted_periods[length(sorted_periods) - (periodsInTestData + 1)]

# Build the 2-row table
period_table = tibble(
  Training = c(min_period, T1),
  Testing   = c(T2, max_period)
)

period_table

```

## Dataset splitting by period
```{r}
# ------------------------------------------------------------
# Split into training and test sets
# ------------------------------------------------------------

train_df = df_no_earliest %>%
  filter(periodStart <= T1) %>%
  select(-periodStart)   # drop original date column

test_df = df_no_earliest %>%
  filter(periodStart >= T2) %>%
  select(-periodStart)   # drop original date column


train_df$earliest_period = NULL
test_df$earliest_period = NULL

head(train_df)
head(test_df)


```


```{r}
#Write the training and test dfs to files for reference:
write.csv(train_df, "train_df.csv", row.names = FALSE)
write.csv(test_df,  "test_df.csv",  row.names = FALSE)

```

# Imbalance Ratio
```{r}
print("Imbalance ratio of df:")
df %>%
  count(Target_ConflictFlag) %>%
  mutate(
    proportion = n / sum(n),
    imbalance_ratio = max(n) / min(n)
  )

print("Imbalance ratio of train_df:")
train_df %>%
  count(Target_ConflictFlag) %>%
  mutate(
    proportion = n / sum(n),
    imbalance_ratio = max(n) / min(n)
  )

print("Imbalance ratio of test_df:")
test_df %>%
  count(Target_ConflictFlag) %>%
  mutate(
    proportion = n / sum(n),
    imbalance_ratio = max(n) / min(n)
  )
```


```{r, echo = FALSE}
#Install Catboost (requires rlang 1.2.0+ and rtools 4.5+)


#install.packages("devtools") 
devtools::install_github("catboost/catboost", subdir = "catboost/R-package")
library(catboost)

library(dplyr)
```

# Model Training
```{r, echo=FALSE}
# ------------------------------------------------------------
# 6. train a catboost model based on country and place forming the composite key. 
# ------------------------------------------------------------


# Target column is now the binary conflict flag

print(cat("Target column: ", target_colname))

cat_features <- c("country", "place")

# Prepare CatBoost training pool
train_pool <- catboost.load_pool(
  data  = train_df %>% select(-all_of(target_colname)),
  label = train_df[[target_colname]],
  cat_features = cat_features
)

# Train the global CatBoost model (classification)
params <- list(
  loss_function = "Logloss",     # <-- IMPORTANT: binary classification
  eval_metric   = "AUC",         # optional but recommended
  iterations    = 500,
  depth         = 8,
  learning_rate = 0.05,
  has_time = TRUE
)

model_catBoost <- catboost.train(train_pool, params = params)

```



# Evaluation of Model
```{r}
# ------------------------------------------------------------
# 4. Compute training evaluation metrics
# ------------------------------------------------------------

test_pool <- catboost.load_pool(
  data  = test_df %>% select(-all_of(target_colname)),
  label = test_df[[target_colname]],
  cat_features = cat_features
)

# ------------------------------------------------------------
# Generate predictions
# ------------------------------------------------------------

# ------------------------------------------------------------
# 4. Compute evaluation metrics for binary classification
# ------------------------------------------------------------

test_pool <- catboost.load_pool(
  data  = test_df %>% select(-all_of(target_colname)),
  label = test_df[[target_colname]],
  cat_features = cat_features
)

# ------------------------------------------------------------
# Generate predictions (probabilities)
# ------------------------------------------------------------

pred_probs <- catboost.predict(model_catBoost, test_pool)

# Actual binary labels
actuals <- test_df[[target_colname]]

# ------------------------------------------------------------
# Convert probabilities to class predictions
# ------------------------------------------------------------

threshold <- 0.5
pred_classes <- ifelse(pred_probs >= threshold, 1, 0)

# ------------------------------------------------------------
# Classification metrics
# ------------------------------------------------------------

# AUC
auc <- pROC::auc(actuals, pred_probs)

# Accuracy
accuracy <- mean(pred_classes == actuals)

# Precision
precision <- sum(pred_classes == 1 & actuals == 1) /
             sum(pred_classes == 1)

# Recall (Sensitivity)
recall <- sum(pred_classes == 1 & actuals == 1) /
          sum(actuals == 1)

# F1 Score
f1 <- 2 * (precision * recall) / (precision + recall)

# Confusion Matrix
conf_matrix <- table(Predicted = pred_classes, Actual = actuals)

# ------------------------------------------------------------
# Print results
# ------------------------------------------------------------

cat("AUC:        ", auc, "\n")
cat("Accuracy:   ", accuracy, "\n")
cat("Precision:  ", precision, "\n")
cat("Recall:     ", recall, "\n")
cat("F1 Score:   ", f1, "\n")
cat("Confusion Matrix:\n")
print(conf_matrix)


```


